Automatically Proving Equivalence by Type-Safe Reflection

### 2017-03-29 paper-CICM-17/content/1-introduction.tex
paper almost ready, still need to proofread a lot

Proofs assistants like Coq(ref), Agda(ref) and Idris(ref) are based on Intensional Type Theories, that contain two notions of equality. One is the equality that can be manipulated in the language, called the propositional equality. This equality corresponds to the equality found in mathematics : this is a proposition that can be assumed, negated, proved or disproved. Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parametrised by the type $a$) indexed over two elements of $a$, giving $Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $(Id_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

Proving that one term is equal to another is quite common
in formal certification. This is an everyday routine when working in a dependently typed language, and especially when intensively indexing types over values in order to capture some logical properties, as we will show in the following subsection.

@@

We revisit an example from a previous work of Edwin Brady~\cite{bradytfp07} which shows how proof obligations may arise when a type is indexed by natural numbers. In this development, the goal is to implement a formally verified library of binary numbers. In order to ensure the correctness of the methods that will be developed, we define the types $Bit$ and $Binary$ as indexed over the value they represent (expressed as a natural number), following a correct-by-construction approach.

@@

The proof that we have written by hand only uses the existence of neutral elements, and the associativity and commutativity of $+$ on $Nat$. Thus, we're rewriting a term by using the properties of a commutative monoid. We would like to let a generic prover for commutative monoid find a proof on its own.

@@

Ring solvers are already implemented for various proof
assistants, including Coq~\cite{Coq2005} and Agda\footnote{\url{http://wiki.portal.chalmers.se/agda/\%22?n=Libraries.UsingTheRingSolver}}. 
In this paper, we describe a certified
implementation\footnote{The implementation of our hierarchy of tactics can be found online at \url{https://github.com/FranckS/RingIdris}} of an automatic prover for equalities in a \emph{hierarchy} of algebraic
structures, including Monoid, Groups and Rings (all potentially commutative),
for the Idris language.

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing tactics has been done several times, including for the implementation of a ring solver for the Coq proof assistant, but without the type-safe aspect. We will
compare our approach with other implementations in Section~\ref{sect:AlternativeApproaches}.

>>>>

Proofs assistants like Coq~\cite{BertotC04}, Agda(ref) and Idris~\cite{brady2013idris} are based on Intensional Type Theories, that contain two notions of equality. One is the equality that can be manipulated in the language, called the propositional equality. This equality corresponds to the equality found in mathematics : this is a proposition that can be assumed, negated, proved or disproved. Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parametrised by the type $a$) indexed over two elements of $a$, giving $Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $(Id_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

Proving that one term is equal to another is quite common
in formal certification. This is even an everyday routine when working in a dependently typed language, and especially when intensively indexing types over values in order to capture some logical properties, as we will show in the following subsection.

@@

We revisit an example from a previous work of Edwin Brady~\cite{DBLP:conf/plpv/Brady13} which shows how proof obligations may arise when a type is indexed by natural numbers. In this development, the goal is to implement a formally verified library of binary numbers. In order to ensure the correctness of the methods that will be developed, we define the types $Bit$ and $Binary$ as indexed over the value they represent (expressed as a natural number), following a correct-by-construction approach.

@@

The proof that we have written by hand only uses the existence of a neutral element, and the associativity and commutativity of $+$ on $Nat$. Thus, we're rewriting a term by using the properties of a commutative monoid. We would like to let a generic prover for commutative monoid find a proof on its own.

@@

Prover for some algebraic structures have already been implemented for various proof assistants, including Coq~\cite{Coq2005} and Agda\footnote{\url{http://wiki.portal.chalmers.se/agda/\%22?n=Libraries.UsingTheRingSolver}} ring prover. In this paper, we describe a certified
implementation\footnote{The implementation of our hierarchy of tactics can be found online at \url{https://github.com/FranckS/RingIdris}} of an automatic prover for equalities in a \emph{hierarchy} of algebraic
structures, including Monoid, Groups and Rings (all potentially commutative),
for the Idris language.

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing tactics has been done several times, including for the implementation of a ring solver for the Coq proof assistant, but without the type-safe and correct-by-construction aspects. We will
compare our approach with other implementations in Section~\ref{sect:relatedWork}.

### 2017-03-29 paper-CICM-17/content/3-hierarchyOfTactics.tex
paper almost ready, still need to proofread a lot

What we want to build is a prover able to prover not only equalities for $Nat$, $List$,or nay specific type but for various datatypes, and for various properties. With some abstraction, we can solve the problem of generating proof of equality for many datatypes like natural numbers, lists, matrix, and many others \emph{all at once} by implementing a very generic hierarchy of provers that can prove equalities of terms for Semi-Groups, Monoids, Commutative-Monoids, Groups, Commutative-Groups, Semi-Rings and Rings. These tactics will be usable on any type that satisfies the properties of one of these algebraic structure. The properties of a given algebraic algebraic structure will be expressed in a corresponding interface (sometimes called typeclass in other progamming languages). This interface will extend the interface from which it naturally inherits. For example, the $Group$ interface will extend the $Monoid$ one. We will end up with a hierarchy of interfaces and will write one tactic for each of these interface. All the tactics that we will implement (the ring prover, the group prover, etc) will be able to work on any type, being given that an implementation of the corresponding interface is provided.

>>>>

What we want to build is a prover able to prover not only equalities for $Nat$, $List$, or any specific type but for various datatypes, and for various properties. With some abstraction, we can solve the problem of generating proof of equality for many datatypes like natural numbers, lists, matrix, and many others \emph{all at once} by implementing a very generic hierarchy of provers that can prove equalities of terms for Semi-Groups, Monoids, Commutative-Monoids, Groups, Commutative-Groups, Semi-Rings and Rings. These tactics will be usable on any type that satisfies the properties of one of these algebraic structure. The properties of a given algebraic algebraic structure will be expressed in a corresponding interface (sometimes called typeclass in other progamming languages). This interface will extend the interface from which it naturally inherits. For example, the $Group$ interface will extend the $Monoid$ one. We will end up with a hierarchy of interfaces and will write one tactic for each of these interface. All the tactics that we will implement (the ring prover, the group prover, etc) will be able to work on any type, being given that an implementation of the corresponding interface is provided.

### 2017-03-29 paper-CICM-17/content/4-relatedWork.tex
paper almost ready, still need to proofread a lot

As mentioned, Coq's implementation of a ring prover was also using proof-by-reflection techniques, but without the guarantees obtained with our type-safe reflection mechanism, and without the correct-by-construction approach : they have first defined the normalisation of terms, and afterwards they've proven the correctness of the normalisation with the external lemma :
$\forall\ (e1\ e2\ :\ Expr),\ beq_{Expr}\ (norm\ e1)\ (norm\ e2)\ =\ true\ \rightarrow\ reify\ e1\ \simeq\ reify\ e2$ which therefore needs a $reify$ function that we do not need in our development.
Our automatic reflection was written with Idris reflection mechanism which allows to do pattern matching on ssyntax, and their automatic reflection is programmed in Ltac~\cite{DelahayeLTac} : a proof dedicated and untyped meta-language for the writing of automations. \\

Proof automations have been investigated for various proof assistants and programming languages. In Coq, appart from the $ring$ prover~\cite{Coq2005}, there's also the $omega$ solver~\cite{Cregut04}, which solves a goal in Presburger arithmetic (i.e. a universally quantified formula made of equations and inequations), and a Field~\cite{DelahayeField} decision procedure for real numbers, which plugs to Coq's ring prover after simplification of the multiplicative inverses. Various automations have also been done in Agda~\cite{Lindblad04}. \\

>>>>

As mentioned, Coq's implementation of a ring prover was also using proof-by-reflection techniques, but without the guarantees obtained with our type-safe reflection mechanism, and without the correct-by-construction approach : they have first defined the normalisation of terms, and afterwards they've proven the correctness of the normalisation with the external lemma :
$\forall\ (e1\ e2\ :\ Expr),\ beq_{Expr}\ (norm\ e1)\ (norm\ e2)\ =\ true\ \rightarrow\ reify\ e1\ \simeq\ reify\ e2$ which therefore needs a $reify$ function that we do not need in our development.
Our automatic reflection was written with Idris reflection mechanism which allows to do pattern matching on syntax, and their automatic reflection is programmed in Ltac~\cite{DelahayeLTac} : a proof dedicated and untyped meta-language for the writing of automations. \\

Proof automations have been investigated for various proof assistants and programming languages. In Coq, appart from the $ring$ prover~\cite{Coq2005}, there's also the $omega$ solver~\cite{Cregut04}, which solves a goal in Presburger arithmetic (i.e. a universally quantified formula made of equations and inequations), and a Field~\cite{DelahayeField} decision procedure for real numbers, which plugs to Coq's ring prover after simplification of the multiplicative inverses. Various automations have also been done in Agda~\cite{DBLP:conf/mpc/KokkeS15} ~\cite{Lindblad04}. \\
